---
title: "Step 4: Retention Time Alignment Visualization"
author: "xcmsVis Package"
date: today
format:
  html:
    toc: true
    toc-location: left
    code-fold: false
    theme: cosmo
vignette: >
  %\VignetteIndexEntry{Step 4: Retention Time Alignment Visualization}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  warning = FALSE,
  message = FALSE
)
```

# Introduction

This vignette covers the **fourth step** in the XCMS metabolomics workflow: **retention time alignment**. After peak correspondence, these functions help you:

- Visualize retention time corrections across samples
- Assess alignment quality
- Compare different alignment methods
- Examine alignment model parameters (LamaParama)

## XCMS Workflow Context

```
┌───────────────────────────────┐
│ 1. Raw Data Visualization     │
│ 2. Peak Detection             │
│ 3. Peak Correspondence        │
├───────────────────────────────┤
│ 4. RT ALIGNMENT               │ ← YOU ARE HERE
├───────────────────────────────┤
│ 5. Feature Grouping           │
└───────────────────────────────┘
```

## What is Retention Time Alignment?

Retention time alignment (also called retention time correction) adjusts for systematic shifts in retention time between samples. This improves feature detection by ensuring that the same compound elutes at consistent times across all samples.

## Functions Covered

- **`gplotAdjustedRtime()`**: Visualize retention time corrections for any alignment method
- **`gplot(LamaParama)`**: Visualize the LamaParama alignment model specifically

# Setup

```{r libraries}
library(xcms)
library(xcmsVis)
library(ggplot2)
library(plotly)
library(faahKO)
library(MsExperiment)
library(BiocParallel)
```

# Data Preparation

We'll use the faahKO package data:

```{r load_data}
# Get example CDF files
cdf_files <- dir(system.file("cdf", package = "faahKO"),
                  recursive = TRUE, full.names = TRUE)

# Using 6 samples (3 from each group)
cdf_files <- cdf_files[c(1:3, 7:9)]

# Load data as XcmsExperiment
xdata_raw <- readMsExperiment(
  spectraFiles = cdf_files,
  BPPARAM = SerialParam()
)

# Add sample metadata
sample_group <- rep(c("KO", "WT"), each = 3)
sampleData(xdata_raw)$sample_name <- basename(cdf_files)
sampleData(xdata_raw)$sample_group <- sample_group
```

## Peak Detection and Correspondence

```{r peak_detection}
# Peak detection
cwp <- CentWaveParam(
  peakwidth = c(20, 80),
  ppm = 25
)
xdata_peaks <- findChromPeaks(xdata_raw, param = cwp, BPPARAM = SerialParam())

# Initial correspondence (peak grouping)
sample_data <- sampleData(xdata_peaks)

pdp <- PeakDensityParam(
  sampleGroups = sample_data$sample_group,
  minFraction = 0.4,
  bw = 30
)
xdata_grouped <- groupChromPeaks(xdata_peaks, param = pdp)
```

# Part 1: General Alignment Visualization

## gplotAdjustedRtime(): PeakGroups Method

### Basic Workflow

```{r peakgroups_alignment}
# Work with a copy of the grouped data
xdata <- xdata_grouped

# Retention time alignment using peak groups
pgp <- PeakGroupsParam(
  minFraction = 0.4,
  smooth = "loess",
  span = 0.2,
  family = "gaussian"
)
xdata <- adjustRtime(xdata, param = pgp)
```

### Creating the Plot

```{r gplot_adjustedrtime}
#| fig-height: 4
#| fig-width: 5
#| fig-alt: "ggplot2 version of retention time adjustment plot."
p <- gplotAdjustedRtime(xdata, color_by = sample_group)
print(p)
```

The plot shows:

- **Lines**: One per sample, showing RT deviation across the chromatographic run
- **Grey circles**: Individual peaks used for alignment
- **Grey dashed lines**: Connect peaks from the same feature across samples

### Customizing the Plot

```{r custom_plot}
#| fig-height: 5
#| fig-width: 7
#| fig-alt: "Customized retention time adjustment plot."
p +
  labs(
    title = "Retention Time Correction - faahKO Dataset",
    x = "Adjusted Retention Time (s)",
    y = "RT Adjustment (s)",
    color = "Sample Group"
  ) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "top")
```

### Interactive Visualization

```{r interactive_rt}
#| fig-alt: "Interactive plotly version of RT adjustment plot."
p_interactive <- ggplotly(p, tooltip = "text")
p_interactive
```

## Advanced Use Cases

### With filterFile()

```{r with_filter}
# Work with a fresh copy from grouped data
xdata_filtered <- xdata_grouped

# Filter to specific samples (files 2-5)
xdata_filtered <- filterFile(xdata_filtered, c(2:5))

# Filtering removes correspondence - need to re-group with filtered samples
sample_data_filtered <- sampleData(xdata_filtered)
pdp_filtered <- PeakDensityParam(
  sampleGroups = sample_data_filtered$sample_group,
  minFraction = 0.4,
  bw = 30
)
xdata_filtered <- groupChromPeaks(xdata_filtered, param = pdp_filtered)

# Align filtered data
pgp_filter <- PeakGroupsParam(minFraction = 0.4)
xdata_filtered <- adjustRtime(xdata_filtered, param = pgp_filter)
```

```{r plot_filtered}
#| fig-height: 4
#| fig-width: 5
#| fig-alt: "RT adjustment plot for filtered dataset."
gplotAdjustedRtime(xdata_filtered, color_by = sample_group)
```

### With subset Parameter

You can use only specific samples for alignment calculation while keeping all samples in the dataset:

```{r with_subset}
# Work with a fresh copy of grouped data
xdata_subset <- xdata_grouped

# Use subset parameter to align using only specific samples
pgp_subset <- PeakGroupsParam(
  minFraction = 0.4,
  smooth = "loess",
  span = 0.2,
  family = "gaussian",
  subset = c(1, 2, 3, 5)  # Exclude sample 4 from alignment calculation
)
xdata_subset <- adjustRtime(xdata_subset, param = pgp_subset)
```

```{r plot_subset}
#| fig-height: 4
#| fig-width: 5
#| fig-alt: "RT adjustment plot with subset alignment."
gplotAdjustedRtime(xdata_subset, color_by = sample_group)
```

# Part 2: LamaParama Alignment Visualization

## What is LamaParama?

LamaParama (Landmark-based Alignment Parameters) enables alignment of LC-MS datasets to external reference data. This is particularly useful when:

- Comparing samples across different laboratories
- Aligning data collected at different time points
- Matching experimental data to a reference dataset (e.g., from QC samples)

The method works by:
1. Defining **landmarks** (reference features with known m/z and RT)
2. Matching chromatographic peaks from your data to these landmarks
3. Fitting a model (loess, polynomial, etc.) based on matched peaks
4. Applying RT corrections to minimize differences between datasets

## Example: Aligning to Reference Dataset

Following the XCMS vignette approach, we'll align a test dataset to a reference:

```{r lama_setup}
# Load reference dataset (contains QC samples)
ref <- loadXcmsData("xmse")

# Load test dataset (to be aligned)
tst <- loadXcmsData("faahko_sub2")
```

### Extract Landmarks from Reference

Landmarks should be features consistently present across samples. Here we use features found in all QC samples:

```{r lama_landmarks}
# Filter to features present in all QC samples
f <- sampleData(ref)$sample_type
f[f != "QC"] <- NA
ref_filtered <- filterFeatures(ref, PercentMissingFilter(threshold = 0, f = f))

# Extract landmark m/z and RT values
ref_mz_rt <- featureDefinitions(ref_filtered)[, c("mzmed", "rtmed")]
cat("Number of landmarks extracted:", nrow(ref_mz_rt), "\n")
head(ref_mz_rt)
```

### Perform Alignment

```{r lama_alignment}
# Create LamaParama object with landmarks
lama_param <- LamaParama(
  lamas = ref_mz_rt,
  method = "loess",
  span = 0.5,
  outlierTolerance = 3,
  zeroWeight = 10,
  ppm = 20,
  tolerance = 0,
  toleranceRt = 20
)

# Apply alignment to test dataset
tst_adjusted <- adjustRtime(tst, param = lama_param)
tst_adjusted <- applyAdjustedRtime(tst_adjusted)
```

## Visualizing LamaParama Results

### Extract LamaParama Object

```{r extract_lama}
# Extract LamaParama from process history
proc_hist <- processHistory(tst_adjusted, type = xcms:::.PROCSTEP.RTIME.CORRECTION)
lama_result <- proc_hist[[length(proc_hist)]]@param
```

### Visualize Alignment Model

The `gplot(LamaParama)` function visualizes the RT alignment model for each sample:

- **Points**: Chromatographic peaks from test data matched to reference landmarks
- **Line**: Fitted model showing RT corrections

```{r lama_plot}
#| fig-height: 5
#| fig-width: 5
#| fig-alt: "Landmark-based alignment model for sample 1."
# Plot alignment model for first sample
p <- gplot(lama_result, index = 1)
p
```

The x-axis shows observed retention times (test data), the y-axis shows reference retention times (landmarks). The fitted line shows the correction model applied to align the datasets.

### Multiple Samples

```{r lama_multiple}
#| fig-height: 12
#| fig-width: 5
#| fig-alt: "Alignment models for first 3 samples."
# Create plots for first 3 samples
p1 <- gplot(lama_result, index = 1) + labs(title = "Sample 1")
p2 <- gplot(lama_result, index = 2) + labs(title = "Sample 2")
p3 <- gplot(lama_result, index = 3) + labs(title = "Sample 3")

# Display plots
library(patchwork)
p1 / p2 / p3
```

Each plot shows how well the test sample's peaks match the reference landmarks and the fitted correction model.


### Customization

#### Custom Colors

```{r lama_custom}
p_custom <- gplot(lama_result,
                  index = 1,
                  colPoints = "#E41A1C",
                  colFit = "#377EB8") +
  labs(x = "Observed RT", y = "Reference RT")

p_custom
```

#### With ggplot2 Enhancements

```{r lama_enhanced}
gplot(lama_result, index = 1) +
  ggtitle("Landmark-Based Alignment Model") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.title = element_text(size = 12)
  )
```

#### Interactive

```{r lama_interactive}
p_interactive <- gplot(lama_result, index = 1)
ggplotly(p_interactive)
```

## API Differences: gplot(LamaParama) vs plot(LamaParama)

### Side-by-Side Comparison

```{r lama_comparison}
#| fig-height: 4
#| fig-width: 10
#| fig-alt: "Side-by-side comparison of gplot vs XCMS original plot for LamaParama."
# ggplot2 version
p_gg <- gplot(lama_result, index = 1) +
  labs(title = "gplot(LamaParama) - ggplot2", x = "Observed RT", y = "Reference RT")

# XCMS original
par(mar = c(4, 4, 2, 1))
plot(lama_result, index = 1, main = "plot(LamaParama) - Base R")

# Display ggplot2 version
p_gg
```

### Key Differences

| Aspect | `gplot()` (ggplot2) | `plot()` (XCMS) |
|--------|---------------------|-----------------|
| **Graphics System** | ggplot2 | Base R |
| **Customization** | Use `+ labs()`, `+ theme()`, etc. | Pass parameters to `plot()` |
| **Interactive** | Easy with `ggplotly()` | Requires plotly conversion |
| **Layering** | Can add geoms with `+` | Limited layering |
| **Return Value** | ggplot object | NULL (side effect only) |

## Interpretation

### Good Alignment

A good alignment shows:

- Many matched points across the RT range
- Smooth fitted line (not too wiggly)
- Points closely following the fitted line
- Few outliers

### Potential Issues

Watch for:

- **Few matched points**: May indicate poor peak grouping or too strict tolerance
- **Non-smooth fit**: Could indicate overfitting or poor parameter choice
- **Many outliers**: May suggest RT shifts that are too complex for the model
- **Gaps in coverage**: Some RT regions may not be well-represented

## Visualizing Overall Alignment Result

You can use `gplotAdjustedRtime()` to visualize the overall alignment result after applying LamaParama:

```{r lama_overall}
#| fig-height: 5
#| fig-width: 8
#| fig-alt: "Overall RT adjustment visualization after LamaParama alignment."
sample_data_tst <- sampleData(tst_adjusted)
gplotAdjustedRtime(tst_adjusted, color_by = sample_group) +
  labs(title = "LamaParama Alignment Result")
```

# Summary

## Functions Covered

| Function | Purpose | Input Type |
|----------|---------|------------|
| `gplotAdjustedRtime()` | General RT alignment visualization | XcmsExperiment, XCMSnExp |
| `gplot(LamaParama)` | LamaParama-specific model visualization | LamaParama object |

## Use Cases

- **Quality control**: Verify alignment quality across samples
- **Method comparison**: Compare PeakGroups, Obiwarp, LamaParama
- **Parameter optimization**: Tune alignment parameters
- **Troubleshooting**: Identify samples with problematic alignment

## Next Steps

After aligning retention times, proceed to:

→ **[Step 5: Feature Grouping](05-feature-grouping.html)** - Group features (isotopes, adducts)

# Comparison with Original XCMS

<div class="row">
<div class="col-md-6">

## Original XCMS

```{r original_rt}
#| fig-height: 4
#| fig-width: 5
#| fig-alt: "XCMS plotAdjustedRtime using base R graphics."
sample_data <- sampleData(xdata)

plotAdjustedRtime(
  xdata,
  col = as.factor(sample_data$sample_group),
  peakGroupsCol = "grey"
)
```

</div>
<div class="col-md-6">

## xcmsVis ggplot2

```{r xcmsvis_rt}
#| fig-height: 4
#| fig-width: 5
#| fig-alt: "ggplot2 version with modern aesthetics."
gplotAdjustedRtime(xdata, color_by = sample_group)
```

</div>
</div>

# Session Info

```{r session_info}
sessionInfo()
```
