---
title: "Comparing xcmsVis and XCMS Visualization Approaches"
author: "xcmsVis Package"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Comparing xcmsVis and XCMS Visualization Approaches}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  warning = FALSE,
  message = FALSE
)
```

# Introduction

This vignette demonstrates the differences between the original XCMS plotting functions and their ggplot2-based equivalents in `xcmsVis`. The main advantages of the ggplot2 approach include:

- **Modern aesthetics**: Clean, publication-ready graphics
- **Interactivity**: Easy conversion to interactive plots using `plotly`
- **Consistency**: Uniform styling across all visualizations
- **Flexibility**: Easy customization using standard ggplot2 syntax

# Setup

```{r libraries}
library(xcms)
library(xcmsVis)
library(ggplot2)
library(plotly)
library(faahKO)
library(MsExperiment)
```

# Data Preparation

For this example, we'll use the faahKO package, which contains example LC-MS data from an analysis of fatty acid amide hydrolase (FAAH) knockout mice.

**Note**: All xcmsVis functions work with `XcmsExperiment` (XCMS version 4+) objects. They also maintain backward compatibility with `XCMSnExp` (XCMS version 3) objects.

```{r load_data}
# Get example CDF files from faahKO package
cdf_files <- dir(system.file("cdf", package = "faahKO"),
                  recursive = TRUE, full.names = TRUE)

# For this vignette, we'll use a subset for faster processing
# Using 3 samples from each group
cdf_files <- cdf_files[c(1:3, 7:9)]

# Load data as XcmsExperiment
xdata <- readMsExperiment(spectraFiles = cdf_files)

# Add sample metadata
sample_group <- rep(c("KO", "WT"), each = 3)
sampleData(xdata)$sample_name <- basename(cdf_files)
sampleData(xdata)$sample_group <- sample_group
```

```{r filter_samples}
xdata <- filterFile(xdata, c(2:6))
```


```{r example_workflow}
# Peak detection using CentWave algorithm
cwp <- CentWaveParam(
  peakwidth = c(20, 80),
  ppm = 25
)
xdata <- findChromPeaks(xdata, param = cwp)

# Initial correspondence (peak grouping) before alignment
pdp <- PeakDensityParam(
  sampleGroups = sampleData(xdata)$sample_group,
  minFraction = 0.4,
  bw = 30
)
xdata <- groupChromPeaks(xdata, param = pdp)

# Retention time alignment using peak groups
pgp <- PeakGroupsParam(
  minFraction = 0.4,
  smooth = "loess",
  span = 0.2,
  family = "gaussian",
  subset = c(1,2,3,5) # to test that the function works correctly with subsets
)
xdata <- adjustRtime(xdata, param = pgp)

# Re-do correspondence after alignment
xdata <- groupChromPeaks(xdata, param = pdp)
```

# Comparing plotAdjustedRtime and gplotAdjustedRtime

## Original XCMS Version

The original `plotAdjustedRtime()` function from XCMS uses base R graphics:

```{r original_plot}
# Base R graphics version
plotAdjustedRtime(
  xdata,
  col = as.factor(sampleData(xdata)$sample_group),
  peakGroupsCol = "grey"
)
```

**Characteristics:**
- Uses base R graphics system
- Limited customization options
- Not easily made interactive
- Coloring requires manual color specification

## xcmsVis ggplot2 Version

The `gplotAdjustedRtime()` function provides a modern alternative:

```{r ggplot_version, fig.height=5, fig.width=7}
# ggplot2 version
p <- gplotAdjustedRtime(
  xdata,
  color_by = sample_group
)
print(p)

# Add custom styling
p <- p +
  labs(
    title = "Retention Time Correction",
    x = "Adjusted Retention Time (s)",
    y = "RT Adjustment (s)",
    color = "Sample Group"
  ) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1")

print(p)
```

**Advantages:**
- Modern ggplot2 aesthetics
- Easy to customize using standard ggplot2 syntax
- Automatic legend generation
- Publication-ready output

## Interactive Visualization

One of the key advantages of the ggplot2 approach is easy conversion to interactive plots:

```{r interactive, eval=FALSE}
# Convert to interactive plotly plot
# Note: Interactivity doesn't render in static vignettes,
# but works when run interactively in R
p_interactive <- ggplotly(p, tooltip = "text")
p_interactive
```

The interactive version allows you to:

- Hover over data points to see sample information
- Zoom into regions of interest
- Pan across the plot
- Toggle samples on/off by clicking the legend
- Export as static image

# Detailed Feature Comparison

## Feature-by-Feature Breakdown

| Feature | XCMS `plotAdjustedRtime` | xcmsVis `gplotAdjustedRtime` |
|---------|--------------------------|------------------------------|
| Graphics system | Base R | ggplot2 |
| Interactivity | No | Yes (via plotly) |
| Color by variable | Manual colors only | Automatic by grouping variable |
| Customization | Limited parameters | Full ggplot2 customization |
| Tooltips | No | Yes (sample metadata) |
| Publication quality | Basic | High |
| Legend | Manual | Automatic |
| Theme support | No | Yes |

## Use Cases

### When to use XCMS's plotAdjustedRtime:

- Quick diagnostic checks during analysis
- When base R graphics are preferred
- In existing workflows that rely on base graphics

### When to use xcmsVis's gplotAdjustedRtime:

- Creating publication figures
- Interactive data exploration
- When you need to customize appearance
- When analyzing complex experiments with many sample groups
- For presentations and reports

# Code Comparison Example

Here's a side-by-side comparison of creating a styled plot:

## XCMS Approach

```{r xcms_styled}
# Requires manual color management and layout
colors <- rep(c("#E41A1C", "#377EB8"), each = 3)
plotAdjustedRtime(
  xdata,
  col = colors,
  lwd = 2,
  peakGroupsCol = "#00000040",
  main = "RT Correction"
)
legend(
  "topleft",
  legend = c("KO", "WT"),
  col = c("#E41A1C", "#377EB8"),
  lwd = 2,
  bty = "n"
)
```

## xcmsVis Approach

```{r xcmsvis_styled, fig.height=5, fig.width=7}
# Clean, declarative syntax
gplotAdjustedRtime(xdata, color_by = sample_group) +
  labs(
    title = "RT Correction",
    x = "Adjusted RT (s)",
    y = "RT Adjustment (s)"
  ) +
  theme_minimal() +
  theme(legend.position = "top")
```

# Future Additions

This vignette will be expanded as more plotting functions are added to xcmsVis:

- `gplotQC()` - Quality control plots
- `gplotChromPeaks()` - Chromatographic peak visualization
- `gplotChromPeakDensity()` - Peak density visualization
- `gplotEIC()` - Extracted ion chromatograms
- And more!

# Session Information

```{r session_info}
sessionInfo()
```

# References

- XCMS GitHub Repository: https://github.com/sneumann/xcms
- Original Feature Request: https://github.com/sneumann/xcms/issues/551
- Metabonaut Tutorials: https://github.com/rformassspectrometry/Metabonaut
- ggplot2 Documentation: https://ggplot2.tidyverse.org/
- plotly for R: https://plotly.com/r/
