---
title: "Visualizing Retention Time Adjustment with gplotAdjustedRtime"
author: "xcmsVis Package"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Visualizing Retention Time Adjustment with gplotAdjustedRtime}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  warning = FALSE,
  message = FALSE
)
```

# Introduction

This vignette demonstrates the `gplotAdjustedRtime()` function, which provides a modern ggplot2-based visualization of retention time correction in LC-MS metabolomics data. The function is designed to work with both `XcmsExperiment` (XCMS v4+) and `XCMSnExp` (XCMS v3) objects.

# Setup

```{r libraries}
library(xcms)
library(xcmsVis)
library(ggplot2)
library(plotly)
library(faahKO)
library(MsExperiment)
```

# Data Preparation

We'll use the faahKO package, which contains example LC-MS data from an analysis of fatty acid amide hydrolase (FAAH) knockout mice.

```{r load_data}
# Get example CDF files from faahKO package (load once, reuse throughout)
cdf_files <- dir(system.file("cdf", package = "faahKO"),
                  recursive = TRUE, full.names = TRUE)

# Using 3 samples from each group (6 total)
cdf_files <- cdf_files[c(1:3, 7:9)]

# Load data as XcmsExperiment
xdata_raw <- readMsExperiment(spectraFiles = cdf_files)

# Add sample metadata
sample_group <- rep(c("KO", "WT"), each = 3)
sampleData(xdata_raw)$sample_name <- basename(cdf_files)
sampleData(xdata_raw)$sample_group <- sample_group
```

## Peak Detection (performed once)

```{r peak_detection}
# Peak detection using CentWave algorithm (done once for efficiency)
cwp <- CentWaveParam(
  peakwidth = c(20, 80),
  ppm = 25
)
xdata_peaks <- findChromPeaks(xdata_raw, param = cwp)
```

# Basic Usage

## Complete XCMS Workflow

The proper XCMS workflow for retention time adjustment using peak groups is:

1. Peak detection (already done above)
2. Initial correspondence (grouping)
3. Retention time adjustment
4. Re-do correspondence

```{r xcms_workflow}
# Work with a copy for the basic workflow
xdata <- xdata_peaks

# Initial correspondence (peak grouping) before alignment
pdp <- PeakDensityParam(
  sampleGroups = sampleData(xdata)$sample_group,
  minFraction = 0.4,
  bw = 30
)
xdata <- groupChromPeaks(xdata, param = pdp)

# Retention time alignment using peak groups
pgp <- PeakGroupsParam(
  minFraction = 0.4,
  smooth = "loess",
  span = 0.2,
  family = "gaussian"
)
xdata <- adjustRtime(xdata, param = pgp)

# Re-do correspondence after alignment
xdata <- groupChromPeaks(xdata, param = pdp)
```

## Creating the Plot

:::::: {.columns}
::: {.column width="48%"}
### Original XCMS Version

```{r original_plot, fig.height=4, fig.width=5}
# Base R graphics version
plotAdjustedRtime(
  xdata,
  col = as.factor(sampleData(xdata)$sample_group),
  peakGroupsCol = "grey"
)
```
:::

::: {.column width="4%"}
<!-- spacer -->
:::

::: {.column width="48%"}
### xcmsVis ggplot2 Version

```{r ggplot_version, fig.height=4, fig.width=5}
# ggplot2 version
p <- gplotAdjustedRtime(xdata, color_by = sample_group)
print(p)
```
:::
::::::

The plot shows:
- **Lines**: One per sample, showing RT deviation across the chromatographic run
- **Grey circles**: Individual peaks used for alignment
- **Grey dashed lines**: Connect peaks from the same feature across samples

## Customizing the Plot

Since `gplotAdjustedRtime()` returns a ggplot object, you can easily customize it:

```{r customized_plot, fig.height=5, fig.width=7}
p +
  labs(
    title = "Retention Time Correction - faahKO Dataset",
    x = "Adjusted Retention Time (s)",
    y = "RT Adjustment (s)",
    color = "Sample Group"
  ) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "top")
```

## Interactive Visualization

Convert to an interactive plotly plot:

```{r interactive, eval=FALSE}
# Make it interactive
p_interactive <- ggplotly(p, tooltip = "text")
p_interactive
```

# Advanced Use Cases

## With filterFile()

Sometimes you may want to filter samples before alignment:

```{r with_filter}
# Work with a fresh copy from peak-detected data
xdata_filtered <- xdata_peaks

# Group peaks before filtering
xdata_filtered <- groupChromPeaks(xdata_filtered, param = pdp)

# Filter to specific samples (files 2-5)
xdata_filtered <- filterFile(xdata_filtered, c(2:5))

# Align filtered data
pgp_filter <- PeakGroupsParam(minFraction = 0.4)
xdata_filtered <- adjustRtime(xdata_filtered, param = pgp_filter)
xdata_filtered <- groupChromPeaks(xdata_filtered, param = pdp)
```

:::::: {.columns}
::: {.column width="48%"}
### Original XCMS

```{r original_filtered, fig.height=4, fig.width=5}
plotAdjustedRtime(
  xdata_filtered,
  col = as.factor(sampleData(xdata_filtered)$sample_group),
  peakGroupsCol = "grey"
)
```
:::

::: {.column width="4%"}
<!-- spacer -->
:::

::: {.column width="48%"}
### xcmsVis Version

```{r ggplot_filtered, fig.height=4, fig.width=5}
p_filtered <- gplotAdjustedRtime(xdata_filtered, color_by = sample_group)
print(p_filtered)
```
:::
::::::

## With subset Parameter in PeakGroupsParam

You can use only specific samples for alignment calculation while keeping all samples in the dataset:

```{r with_subset}
# Work with a fresh copy
xdata_subset <- xdata_peaks

# Group peaks
xdata_subset <- groupChromPeaks(xdata_subset, param = pdp)

# Use subset parameter to align using only specific samples
pgp_subset <- PeakGroupsParam(
  minFraction = 0.4,
  smooth = "loess",
  span = 0.2,
  family = "gaussian",
  subset = c(1, 2, 3, 5)  # Exclude sample 4 from alignment calculation
)
xdata_subset <- adjustRtime(xdata_subset, param = pgp_subset)
xdata_subset <- groupChromPeaks(xdata_subset, param = pdp)
```

:::::: {.columns}
::: {.column width="48%"}
### Original XCMS

```{r original_subset, fig.height=4, fig.width=5}
plotAdjustedRtime(
  xdata_subset,
  col = as.factor(sampleData(xdata_subset)$sample_group),
  peakGroupsCol = "grey"
)
```
:::

::: {.column width="4%"}
<!-- spacer -->
:::

::: {.column width="48%"}
### xcmsVis Version

```{r ggplot_subset, fig.height=4, fig.width=5}
p_subset <- gplotAdjustedRtime(xdata_subset, color_by = sample_group)
print(p_subset)
```
:::
::::::

## Without subset Parameter (All Samples)

For comparison, here's alignment using all samples:

```{r without_subset}
# Work with a fresh copy
xdata_no_subset <- xdata_peaks

# Group peaks
xdata_no_subset <- groupChromPeaks(xdata_no_subset, param = pdp)

# Align without subset parameter (use all samples)
pgp_no_subset <- PeakGroupsParam(minFraction = 0.4)
xdata_no_subset <- adjustRtime(xdata_no_subset, param = pgp_no_subset)
xdata_no_subset <- groupChromPeaks(xdata_no_subset, param = pdp)
```

:::::: {.columns}
::: {.column width="48%"}
### Original XCMS

```{r original_no_subset, fig.height=4, fig.width=5}
plotAdjustedRtime(
  xdata_no_subset,
  col = as.factor(sampleData(xdata_no_subset)$sample_group),
  peakGroupsCol = "grey"
)
```
:::

::: {.column width="4%"}
<!-- spacer -->
:::

::: {.column width="48%"}
### xcmsVis Version

```{r ggplot_no_subset, fig.height=4, fig.width=5}
p_no_subset <- gplotAdjustedRtime(xdata_no_subset, color_by = sample_group)
print(p_no_subset)
```
:::
::::::

# Advantages of gplotAdjustedRtime

**Compared to XCMS's `plotAdjustedRtime()`:**

- Modern ggplot2 aesthetics
- Easy customization using standard ggplot2 syntax
- Interactive plotting via plotly
- Automatic legend generation
- Consistent with other ggplot2-based visualizations
- Publication-ready output

# Session Information

```{r session_info}
sessionInfo()
```

# References

- XCMS GitHub Repository: https://github.com/sneumann/xcms
- Original Feature Request: https://github.com/sneumann/xcms/issues/551
- faahKO package: Bioconductor example dataset
- ggplot2 Documentation: https://ggplot2.tidyverse.org/
- plotly for R: https://plotly.com/r/
